The effect of detector nonlinearity on \wfa\ PSF profiles for weak gravitational lensing measurements

### 2016-05-27 ms.tex
Chaz edits based on ref response

Assuming that the dominant contribution to nonlinearity is the varying capacitance of the pixel p-n junction, the correction to the detector signal is well approximated by a quadratic term:
\begin{align}
S(Q)= Q - \beta Q ^2
\label{NL}
\end{align}
Here, $Q$ is the true number of elementary charges collected in the pixel, $S$ is the number inferred from the voltage change at the sense node, and $\beta$ is a constant.  To estimate $\beta$, we compute $Q(V)=C(V)*V/q_e$, where $q_e$ is the elementary charge, and the total capacitance is given by the varying junction capacitance plus a constant $C(V) = C_{\rm jn}(V) + C_{\rm fix}$. The junction capacitance varies as (\citealt{mccaughrean87})
\begin{equation}
C_{\rm jn}(V) \propto (1+V/V_{bi})^{-1/2} \;.
\end{equation}
Here, $V_{bi}$ is the ``built in'' potential of the junction and $V=V_{\rm DSUB}-V_{\rm RESET}+\delta V$, where $V_{\rm DSUB}$ is the constant potential at the diode cathode, $V_{\rm RESET}$ is the initial potential at the anode, and $\delta V$ is the change in anode potential due to accumulated photocharge.  As an example, we \textcolor{blue}{substitute} measurements of a 2.4$\mu$m cutoff H2RG by Finger 2006:\footnote{\url{https://www.eso.org/sci/meetings/2006/neon-2006/Finger.pdf}}
\begin{eqnarray*}
V_{\rm bi} &=& 0.412\ {\mbox V} \\
V_{\rm DSUB} &=& 1\ {\mbox V} \\
V_{\rm RESET} &=& 0.5\ {\mbox V} \\
C_{\rm fix} &=& 17.8\ {\mbox fF} \\
C_{\rm jn}(V=.912\ {\mbox V}) &=& 30\ {\mbox fF}
\end{eqnarray*}
Using these parameters to compute $Q(V)$ for $ 0\le \delta V \le 0.3$ V (corresponding to maximum $Q$=60115) and then inverting to find $S(Q)$, we find it is well-fit (to 0.1\% or better) by Eq. \ref{NL} with $\beta=1.18 \times 10^{-6}$. In practice, HXRG calibrations have included additional polynomial parameters which can reduce residuals, extend the range of valid $Q$, and account for additional nonlinear effects (\citealt{hilbert04}, \citealt{hilbert08}, \citealt{hilbert14}); however, the additional parameters are highly degenerate, resulting in large variances in the fitted values.  For our purposes, it suffices to analyze the shape-distorting effects of nonlinearity using a single parameter, $\beta$, which encapsulates most of the effect. In this paper we chose a nominal value of $\beta_0=5\times 10^{-7}$, which is near the midpoint of the measured range for $\beta$ in \citealt{hilbert14}.

@@

The WFI PSF is undersampled by design. In order to maximize the field of view, detectors in instruments of space missions are usually built with a physical size that results in undersampled images, which fail to satisfy the Nyquist-Shannon criterium for the maximum band limit set by the optical response of the system, and therefore produce aliased images.\footnote{The Nyquist-Shannon criterium states that the sampling interval $p$ must satisfy $p < 1/(2 u_{max})$, where $u_{max}$ is the highest frequency in the signal, in order to avoid aliasing.} In general, it is not possible to recover all the information of a continuous function from a discrete sample of points if the image is aliased, and measurements of astronomical object's properties such as magnitude and shape will be erroneous (\citealt{lauer99a}, \citealt{fruchter11}, \citealt{rhodes07}). Also, in particular, we found that our shape measurement algorithm fails when most light is concentrated in a few pixels.

To overcome this problem in real data, multiple dithered exposures are taken, and then processed by an image combination algorithm (\citealt{lauer99b}, \citealt{fruchter02}, \citealt{bertin06}, \citealt{rowe11}) in order to produce an oversampled image that satisfies the Nyquist-Shannon criterium. To study NL with repeated simulations, however, such external software is computationally expensive.  Instead, we approximate oversampled data by rendering the convolved PSF profile\textemdash including the pixel response at its native scale\textemdash as a \gs\ object at a high resolution,\footnote{\textcolor{blue}{This is done through the method {\tt{galsim.GSObject.drawImage}}. The keyword {\tt{method}} should be set to ``{\tt{no\char`_pixel}}" when calling {\tt{galsim.GSObject.drawImage}}.}} since by default it convolves its argument profile with a boxcar of length {\tt{pixel\char`_scale}}. This extra convolution must be turned off because the effective PSF profile we create already includes this pixel response.} setting the parameter {\tt{scale}} to $p/N$, where $N$ is a positive integer and $p$ is the native pixel scale. 
 Defining $Q\equiv1/p\times u_{\text{max}}$ as the sampling factor ($Q<2$, $Q=2$, and $Q\geq2$ represent under-, critically, and over-sampled images, respectively), we see that in order to produce an over-sampled image ($Q\geq2$), the Nyquist-Shannon criterium implies that $N$ must be given by (see \emph{e.g.}, \citealt{marks09}, \citealt{shapiro13})

@@

Note that the NL effect on the reconstructed image from dithering real exposures will depend on the centroid of the source in the dithered inputs: an exposure where the PSF peaks at a pixel center will have higher NL than an exposure where the peak is spread more evenly over 2-4 pixels. To mimic this behavior in our approximation, \textcolor{blue}{we create a PSF profile whose centroid coordinates are random numbers uniformly distributed over the size of a native detector pixel. We then render the profile at a higher resolution $p/N$ and measure its average size and shape over 100 realizations.} With $N$=3, the variation of the NL correction in the pixels of the oversampled image should be similar to the variations in the 5-8 exposures combined using the WFIRST dither strategy.  In practice, we find that the scatters of the shape measurements over these realizations are negligible relative to the means.

@@

Total and peak flux values at each filter per postage stamp are shown in Table 1. The brightest magnitude used was AB $m=18.3$, based on the peak value of the PSF profile in the Y106 band when rendered into a postage stamp at the native pixel scale and placed at the center of the pixel: $\sim 1.00237\times10^{5}$ electrons, representing about $90\%$ of the typical full well depth in the pixels of the H4RG detectors\textemdash$\ \sim1.1\times10^{5}$ electrons. Note that this peak value is an upper limit and will vary for a slightly offset profile. \textcolor{blue}{As mentioned before, for each postage stamp in our calculations we select a centroid with random coordinates uniformly distributed within the native scale pixel.} In addition, since NL is a function of signal, and since \gs\ conserves total flux when changing pixel scales, the total flux in the higher resolution image must be multiplied by a factor of $N^2$ to preserve the appropriate response per pixel to this effect, correcting for the fact that the new image has a factor of $N^2$ more pixels than the one created at the native scale.

@@

The basic simulation process is summarized by the following steps:  
\begin{itemize}
\item[1.] Create a \wfa\ PSF surface brightness profile with a given flux as prescribed by Eq. \ref{flux}, and convolve the PSF profile with a top-hat pixel response with the size of the native scale of the \wfa\ FPA ($p=0.11$ arcseconds per pixel) to produce an effective PSF. 
\item[2.] Draw (\textcolor{blue}{\emph{i.e.}, render}) the effective PSF profile in a noiseless postage stamp of size $1.5k$ by $1.5k$ ($k=64$) pixels at a higher resolution of of $p/N$, with $N=3$, and multiply the resulting image by $N^2$. In this step, NL is not yet applied, but it will be done so in a later step, so the flux still needs to be adjusted. \textcolor{blue} {In addition, each coordinate of the PSF centroid is a random number drawn from a uniform distribution within the native resolution pixel.}
\item[3.] Create another image of the effective PSF at a higher resolution and with the flux adjusted as in step 2. 
Apply the nonlinearity to the postage stamps according to the transformation $I \mapsto I - \beta I^2$ (\emph{c.f.} Eq. \ref{NL}).
\item[4.] Use the adaptive moments algorithm to measure the shape $e_0=(e_{1,0}, e_{2,0})$ and size $R_0$ of the profile without NL to have as baseline reference.
\item[5.] Measure the shape and size of the object with the sensor effect applied, and calculate the quantities $\Delta e_1$, $\Delta e_2$, and $\frac{\Delta R}{R}$, as defined in Eqs. \ref{delta_e} and \ref{delta_r}. 
\item [6.] Repeat steps 1 to 5 %for different AB magnitudes (fluxes) at the nominal value $\beta_0$ and for different $\beta$ values at a fixed magnitude
, averaging over 100 centroid realizations. The  \textcolor{blue}{ size and ellipticity} values reported will be the mean and the standard deviation over these realizations (see Fig. \ref{f2} and Fig. \ref{f3}).
\item[7.] To study the impact of spatial variability in $\beta$, repeat steps 1 to 6 and assume that the model parameter $\beta$ for each high-resolution pixel can be drawn from a Gaussian distribution with a particular mean $\beta$ and variance $\sigma_{\beta}^2$. Over $M$ realizations (for a fixed centroid), calculate the dispersion values $\sigma_{\Delta e_1}$, $\sigma_{\Delta e_2}$, and $\sigma_{\Delta R/R}$ as a function of $\sigma_{\beta}$. In our simulations we use $M=100$ (see Fig. \ref{f4} and Fig. \ref{f5}).

@@

Under these conditions, Fig. \ref{f3} shows that for the brightest stars in our range, to limit the bias of $\Delta R/R$ ($\Delta e$) in the H158 band to $10^{-4}$  ($4.7\times 10^{-5}$)\textemdash about $10\%$ of the estimated \wf\ error budgets, $\beta$ must be calibrated to $\sim1\%$ ($\sim 2.4\%$) (using $c\sim2\times10^4$ at $m=18.3$ from the upper panel of Fig. \ref{f3}, $c\sim4000$ at $m=18.3$ from the lower right panel of the same figure, and $\beta_0=5\times10^{-7}$). For this error tolerance, if the required calibration cannot be achieved, then one must find a higher star magnitude with a less strict requirement.  Our choice of error tolerance here is somewhat arbitrary and only for illustration, and Eq. \ref{calib} and Fig. \ref{f3} should be used to derive detector requirements once true PSF requirements are known.

@@

\begin{figure}[!h]
\centering
\resizebox{\hsize}{!}{\includegraphics{f4.eps}}
\caption{Impact of spatial variability in the $\beta_0$ parameter, quantified as the standard deviation over $M=100$ realizations (for a fixed centroid) in $\Delta e_1$,  $\Delta e_2$, and $\Delta R/R$  as a function of the standard deviation $ \sigma_{\beta_0}$. 
The source magnitude is fixed at 18.3.}
\label{f4}
\end{figure}
We found that the results in Fig. \ref{f4} are insensitive to the nominal $\beta$ and therefore, in general, $\sigma_{\beta}$  can be treated as an error on $\beta$ estimation. The linear relationships in Fig. \ref{f4} also allows us to plot the slope of each curve for different magnitudes, as was done in Fig. \ref{f3}, and therefore analogous equations to Eq. \ref{calib} can be used to derive requirements on the maximum tolerable dispersion in size and ellipticity errors given a particular set of parameters and requirements. The results are shown in Fig. \ref{f5}. We also fitted a model of the form 
\begin{align}
\frac{\sigma_{\Delta R/ R}} {\sigma_{\beta}} = A_{\text{F}}10^{B_{\text{F}} (m - m_{0})} \ (m_0 =20)
\end{align}
for the PSF size trends in Fig. \ref{f5}, as was done above for Fig. \ref{f3} (\emph{c.f.}, Eq. \ref{power_law}). The fit parameters are presented in Table 2 as well.

>>>>

Assuming that the dominant contribution to nonlinearity is the varying capacitance of the pixel p-n junction, we find that the correction to the detector signal is well-approximated by a quadratic term:
\begin{align}
S(Q)= Q - \beta Q ^2
\label{NL}
\end{align}
Here, $Q$ is the true number of elementary charges collected in the pixel, $S$ is the number inferred from the voltage change at the sense node, and $\beta$ is a constant.  To estimate $\beta$, we compute $Q(V)=C(V)*V/q_e$, where $q_e$ is the elementary charge, and the total capacitance is given by the varying junction capacitance plus a constant $C(V) = C_{\rm jn}(V) + C_{\rm fix}$. The junction capacitance varies as (\citealt{mccaughrean87})
\begin{equation}
C_{\rm jn}(V) \propto (1+V/V_{bi})^{-1/2} \;.
\end{equation}
Here, $V_{bi}$ is the ``built in'' potential of the junction and $V=V_{\rm DSUB}-V_{\rm RESET}+\delta V$, where $V_{\rm DSUB}$ is the constant potential at the diode cathode, $V_{\rm RESET}$ is the initial potential at the anode, and $\delta V$ is the change in anode potential due to accumulated photocharge.  As an example, we \textcolor{blue}{substitute} measurements of a 2.4$\mu$m cutoff H2RG by Finger 2006:\footnote{\url{https://www.eso.org/sci/meetings/2006/neon-2006/Finger.pdf}}
\begin{eqnarray*}
V_{\rm bi} &=& 0.412\ {\mbox V} \\
V_{\rm DSUB} &=& 1\ {\mbox V} \\
V_{\rm RESET} &=& 0.5\ {\mbox V} \\
C_{\rm fix} &=& 17.8\ {\mbox fF} \\
C_{\rm jn}(V=.912\ {\mbox V}) &=& 30\ {\mbox fF}
\end{eqnarray*}
Using these parameters to compute $Q(V)$ for $ 0\le \delta V \le 0.3$ V (corresponding to maximum $Q$=60115) and then inverting to find $S(Q)$, we find it is well-fit (to 0.1\% or better) by Eq. \ref{NL} with $\beta=1.18 \times 10^{-6}$. In practice, HXRG calibrations have included additional polynomial parameters which can reduce residuals, extend the range of valid $Q$, and account for additional nonlinear effects (\citealt{hilbert04}, \citealt{hilbert08}, \citealt{hilbert14}); however, the additional parameters are highly degenerate, resulting in large variances in the fitted values.  For our purposes, it suffices to analyze the shape-distorting effects of nonlinearity using a single parameter, $\beta$, which encapsulates most of the effect. In this paper we chose a nominal value of $\beta_0=5\times 10^{-7}$, which is near the midpoint of the measured range for $\beta$ in \citealt{hilbert14}.

@@

The WFI PSF is undersampled by design. In order to maximize the field of view, detectors in instruments of space missions are usually built with a physical size that results in undersampled images, which fail to satisfy the Nyquist-Shannon criterium for the maximum band limit set by the optical response of the system, and therefore produce aliased images.\footnote{The Nyquist-Shannon criterium states that the sampling interval $p$ must satisfy $p < 1/(2 u_{max})$, where $u_{max}$ is the highest frequency in the signal, in order to avoid aliasing.} In general, it is not possible to recover all the information of a continuous function from a discrete sample of points if the image is aliased, and measurements of astronomical object's properties such as magnitude and shape will be erroneous (\citealt{lauer99a}, \citealt{fruchter11}, \citealt{rhodes07}).  Undersampled data can be fit to a model at the risk of introducing model bias.  In practice, we find that our shape measurement algorithm \citealt{HSM} performs poorly on undersampled images -- when most light is concentrated in one or a few pixels, the image ellipticity is poorly constrained.  Thus we must simulate the effect of NL on oversampled images.

To overcome the problem of undersampling in real data, multiple dithered exposures are taken, and then processed by an image combination algorithm (\citealt{lauer99b}, \citealt{fruchter02}, \citealt{bertin06}, \citealt{rowe11}) in order to produce an oversampled image that satisfies the Nyquist-Shannon criterium. To study NL with repeated simulations, however, such external software is computationally expensive.  Instead, we approximate oversampled data by rendering the convolved PSF profile\textemdash including the pixel response at its native scale\textemdash as a \gs\ object at a high resolution,\footnote{\textcolor{blue}{This is done through the method {\tt{galsim.GSObject.drawImage}}. The keyword {\tt{method}} should be set to ``{\tt{no\char`_pixel}}" when calling {\tt{galsim.GSObject.drawImage}}.} since by default it convolves its argument profile with a boxcar of length {\tt{pixel\char`_scale}}. This extra convolution must be turned off because the effective PSF profile we create already includes this pixel response.} setting the parameter {\tt{scale}} to $p/N$, where $N$ is a positive integer and $p$ is the native pixel scale. 
 Defining $Q\equiv1/p\times u_{\text{max}}$ as the sampling factor ($Q<2$, $Q=2$, and $Q\geq2$ represent under-, critically, and over-sampled images, respectively), we see that in order to produce an over-sampled image ($Q\geq2$), the Nyquist-Shannon criterium implies that $N$ must be given by (see \emph{e.g.}, \citealt{marks09}, \citealt{shapiro13})

@@

Note that the NL effect on the reconstructed image from dithering real exposures will depend on the centroid of the source in the dithered inputs: an exposure where the PSF peaks at a pixel center will have higher NL than an exposure where the peak is spread more evenly over 2-4 pixels. To mimic this behavior in our approximation, \textcolor{blue}{we create a PSF profile whose centroid coordinates are random numbers uniformly distributed over the size of a native detector pixel. We then render the profile at a higher resolution $p/N$ and measure its average size and shape over 100 realizations.} With $N$=3, the variation of the NL correction in the pixels of the oversampled image should be similar to the variations in the 5-8 exposures combined using the WFIRST dither strategy.  In practice, we find that the scatters of the shape measurements over these realizations are negligible relative to the means; therefore, we expect that the effect of NL on the shape of an oversampled image is insensitive to the precise dither pattern.

@@

Total and peak flux values at each filter per postage stamp are shown in Table 1. The peaks are upper limits estimated by assuming the PSF centroid is centered on a pixel.  
The brightest magnitude used was AB $m=18.3$, based on the peak value of the PSF profile in the Y106 band when rendered into a postage stamp at the native pixel scale and placed at the center of the pixel: $\sim 1.00237\times10^{5}$ electrons, representing about $90\%$ of the typical full well depth in the pixels of the H4RG detectors\textemdash$\ \sim1.1\times10^{5}$ electrons.
\textcolor{blue}{Recall that for each postage stamp, we select a centroid with coordinates randomly chosen from a uniform distributed across the native scale pixel.} In addition, since NL is a function of signal, and since \gs\ conserves total flux when changing pixel scales, the total flux in the higher resolution image must be multiplied by a factor of $N^2$ to preserve the appropriate response per pixel to this effect, correcting for the fact that the new image has a factor of $N^2$ more pixels than the one created at the native scale.

@@

The basic simulation process is summarized by the following steps:  
\begin{itemize}
\item[1.] Create a \wfa\ PSF surface brightness profile with a given flux as prescribed by Eq. \ref{flux}, and convolve the PSF profile with a top-hat pixel response with the size of the native scale of the \wfa\ FPA ($p=0.11$ arcseconds per pixel) to produce an effective PSF. 
\item[2.] \textcolor{blue}{Sample the effective PSF profile on a 2D grid to draw} a noiseless postage stamp of size $1.5k$ by $1.5k$ ($k=64$) pixels at a higher resolution of of $p/N$, with $N=3$, and multiply the resulting image by $N^2$. In this step, NL is not yet applied, but it will be done so in a later step, so the flux still needs to be adjusted. 
\textcolor{blue} {ÒThe centroid of the profile is randomly selected from a uniform 2D distribution within the size of a native pixel.}
\item[3.] Create another image of the effective PSF at a higher resolution and with the flux adjusted as in step 2. 
Apply the nonlinearity to the postage stamps according to the transformation $I \mapsto I - \beta I^2$ (\emph{c.f.} Eq. \ref{NL}).
\item[4.] Use the adaptive moments algorithm to measure the shape $e_0=(e_{1,0}, e_{2,0})$ and size $R_0$ of the profile without NL to have as baseline reference.
\item[5.] Measure the shape and size of the object with the sensor effect applied, and calculate the quantities $\Delta e_1$, $\Delta e_2$, and $\frac{\Delta R}{R}$, as defined in Eqs. \ref{delta_e} and \ref{delta_r}. 
\item [6.] Repeat steps 1 to 5 %for different AB magnitudes (fluxes) at the nominal value $\beta_0$ and for different $\beta$ values at a fixed magnitude
, averaging over 100 centroid realizations. The  \textcolor{blue}{ size and ellipticity} values reported will be the mean and the standard deviation over these realizations (see Fig. \ref{f2} and Fig. \ref{f3}).
\item[7.] To study the impact of spatial variability in $\beta$, repeat steps 1 to 6 and assume that the model parameter $\beta$ for each high-resolution pixel can be drawn from a Gaussian distribution with a particular mean $\beta$ and variance $\sigma_{\beta}^2$. Over $M$ realizations (for a fixed centroid), calculate the dispersion values $\sigma_{\Delta e_1}$, $\sigma_{\Delta e_2}$, and $\sigma_{\Delta R/R}$ as a function of $\sigma_{\beta}$. In our simulations we use $M=100$ (see Fig. \ref{f4} and Fig. \ref{f5}).

@@

Under these conditions, Fig. \ref{f3} shows that for the brightest stars in our range, to limit the bias of $\Delta R/R$ ($\Delta e$) in the H158 band to $10^{-4}$  ($4.7\times 10^{-5}$)\textemdash about $10\%$ of the estimated \wf\ error budgets, $\beta$ must be calibrated to $\sim1\%$ ($\sim 2.4\%$) (using $c\sim2\times10^4$ at $m=18.3$ from the upper panel of Fig. \ref{f3}, $c\sim4000$ at $m=18.3$ from the lower right panel of the same figure, and $\beta_0=5\times10^{-7}$). 
Alternatively, this calculation can be used to convert between NL calibration requirements, shape measurement error requirements, and the minimum useable star magnitude: given any two of these, we can obtain the third.  E.g., if NL calibration precision reaches some practical limit, then given a tolerance on shape measurement error, one can find the star magnitude for which the shape measurement bias due to NL matches the tolerance (brighter stars cannot be used to measure the PSF).  Our choice of error tolerance here is only for illustration, and Eq. \ref{calib} and Fig. \ref{f3} should be used to derive detector requirements once true PSF requirements are known.

@@

\textcolor{red}{\textbf{IS THERE ANY BIAS IN R, e, DUE TO RANDOM BETA, OR DOES IT MATCH THE NON-RANDOM CASE??}}

\begin{figure}[!h]
\centering
\resizebox{\hsize}{!}{\includegraphics{f4.eps}}
\caption{Impact of spatial variability in the $\beta_0$ parameter, quantified as the standard deviation over $M=100$ realizations (for a fixed centroid) in $\Delta e_1$,  $\Delta e_2$, and $\Delta R/R$  as a function of the standard deviation $ \sigma_{\beta_0}$. 
The source magnitude is fixed at 18.3.}
\label{f4}
\end{figure}
We found that the results in Fig. \ref{f4} are insensitive to the nominal $\beta$ and therefore, in general, $\sigma_{\beta}$  can be treated as an error on $\beta$ estimation. The linear relationships in Fig. \ref{f4} also allows us to plot the slope of each curve for different magnitudes, as was done in Fig. \ref{f3}, and therefore analogous equations to Eq. \ref{calib} can be used to convert between the dispersion in errors on $\beta$ and the dispersions of errors in size and ellipticity measurements.  This computation may also be useful for deriving calibration requirements, although weak lensing analyses are much more sensitive to shape measurement biases than to small random errors.  The results are shown in Fig. \ref{f5}. We also fitted a model of the form 
\begin{align}
\frac{\sigma_{\Delta R/ R}} {\sigma_{\beta}} = A_{\text{F}}10^{B_{\text{F}} (m - m_{0})} \ (m_0 =20)
\end{align}
for the PSF size trends in Fig. \ref{f5}, as was done above for Fig. \ref{f3} (\emph{c.f.}, Eq. \ref{power_law}). The fit parameters are presented in Table 2 as well.
